Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-01-31 paper.tex
enumeration section

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}. From Fig.~\ref{graphicalModel}, the MAP probability of $(\mathcal{D},\theta)$ is
\begin{equation}
  \probability[\mathcal{D},\theta|X]\propto \probability[\mathcal{D},\theta] \prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]\label{probabilityDSL}
\end{equation}
If we had a $(\mathcal{D},\theta)$ maximizing Eq.~\ref{probabilityDSL}, then we could recover the most likely program for task $x$ by maximizing $\probability[x|p] \probability[p|\mathcal{D},\theta]$.
Through this lens we now take as our goal to maximize Eq.~\ref{probabilityDSL}.
But even \emph{evaluating} Eq.~\ref{probabilityDSL} is intractable because it involves summing over the infinite set of all possible programs. In general, programs are hard-won: finding even a single program that explains a given observation presents a daunting combinatorial search problem.

With this fact in mind, we will instead maximize the following tractable lower bound on Eq.~\ref{probabilityDSL},
which we call $J$:
\begin{equation}
J = \probability[\mathcal{D},\theta]\prod_{x\in X}\log \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]\label{lowerBound}
\end{equation}
This lower bound depends on sets of programs, $\left\{\mathcal{F}_x \right\}_{x\in X}$:

\noindent\textbf{Definition.} The \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is the set of programs discovered by \system where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

We maximize $J$ by alternatingly maximizing it w.r.t. the DSL and the frontiers:
\\\noindent \textbf{Program Synthesis: Maximizing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means attempting to synthesize new programs for each of the tasks.
So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{DSL Induction: Maximizing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.  Section~\ref{grammarInductionSection} explains this step of the algorithm.

In practice, synthesizing programs is extremely difficult because
of how large the search space is. Learning the DSL eases the difficulty of synthesis
by exposing a domain-specific basis for constructing programs.
Another complementary means of meeting search is to learn a
bottom-up \emph{recognition model}

\subsection{Synthesizing Programs from a DSL}\label{explorationSection}

Describe the enumerator, depth first search on the random choices made by the generative model with iterative deepening

Limitation: has a hard time learning arbitrary constants. Solution: solve those constants using specialized solvers.

@@

\begin{algorithm}[tb]
   \caption{Grammar Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$, AIC coefficient $a$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $\log \probability[\mathcal{D}]\stackrel{+}{ = } -\lambda\sum_{p\in \mathcal{D}} \text{size}(p)$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - a|\mathcal{D}|$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ a subtree of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \IF{$\text{score}(\mathcal{D}') > \text{score}(\mathcal{D})$}
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ELSE
     \STATE\textbf{return} $\mathcal{D},\theta^*(\mathcal{D})$
     \ENDIF
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}
\section{Program Representation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.

@@

\subsection{Symbolic Regression}
We show how to use \system to infer programs containing both discrete
structure and continuous parameters. The high-level idea is to synthesize programs with unspecified-real-valued parameters, and to fit those parameters using gradient descent.
Concretely, we ask the algorithm to
solve a set of 1000 symbolic regression problems, each a polynomial of
degree 0, 1, or 2, where our observations $x$ take the form of $N$
input/output examples, which we write as $x = \left\{(i_n,o_n)
\right\}_{n\leq N}$. For example, one task is to infer a program
calculating $3x + 2$, and the observations are the input-output
examples $\left\{(-1,-1),(0,2),(1,5) \right\}$.

@@

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} $\lambda \text{var}.$ body
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p:\alpha\to\cdots\to\beta \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if canUnify}(\tau,\beta) \} $

   \STATE Sample $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.35cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Let $e:\alpha_1\to\alpha_2\to\cdots\to \alpha_K\to\beta$. Unify $\tau$ with $\beta$.
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} $e\;a_1\; a_2\; \cdots\; a_K$
 %  \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

>>>>

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}. We wish to maximize the MAP probability of $\mathcal{D}$:
\begin{equation*}
  \probability[\mathcal{D}|X]\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{equation*}
In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
Conventional NLP approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
\begin{align}
  A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

If we had a $(\mathcal{D},\theta)$ maximizing Eq.~\ref{AIC}, then we could recover the most likely program for task $x$ by maximizing $\probability[x|p] \probability[p|\mathcal{D},\theta]$.
Through this lens we now take as our goal to maximize Eq.~\ref{AIC}.
But even \emph{evaluating} Eq.~\ref{AIC} is intractable because it involves summing over the infinite set of all possible programs. In general, programs are hard-won: finding even a single program that explains a given observation presents a daunting combinatorial search problem.
With this fact in mind, we will instead maximize the following tractable lower bound on Eq.~\ref{AIC},
which we call $J$:
\begin{equation}
J = \log \probability[\mathcal{D},\theta] + \sum_{x\in X}\log \sum_{p\in \mathcal{F}_x} \probability[x|p]\probability[p|\mathcal{D},\theta]\label{lowerBound}
\end{equation}
This lower bound depends on sets of programs, $\left\{\mathcal{F}_x \right\}_{x\in X}$:

\noindent\textbf{Definition.} The \emph{frontier of task $x$}, written $\mathcal{F}_x$,
is a set of programs where $\probability[x|p] > 0$ for all $p\in \mathcal{F}_x$.

We maximize $J$ by alternatingly maximizing it w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.

Searching for programs is extremely difficult because
of how large the search space is. We ease the difficulty of the search by learning a neural recognition model:

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
frontier $\mathcal{F}_x$ by searching for programs maximizing
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta_x$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{ritchie2016deep} for the generative model in
Fig.~\ref{graphicalModel}.

%% Learning the DSL eases the difficulty of synthesis
%% by exposing a domain-specific basis for constructing programs.
%% Another complementary means of meeting search is to learn a
%% bottom-up \emph{recognition model}

\subsection{Searching for Programs}\label{explorationSection}

Now our goal is to search for programs that solve the tasks.  In this
work we use the simple search strategy of enumerating programs from
the DSL  in decreasing order of their probability,
and then checking if an enumerated program $p$ assigns positive
probability to a task ($\probability[x|p] > 0$); if so, we include $p$ in
the frontier $\mathcal{F}_x$.

To make this concrete we need to define what programs actually are and
what form $\probability[p |\mathcal{D},\theta]$ takes.
In this work, we represent programs as \emph{polymorphicly-type $\lambda$-calculus expressions}.
$\lambda$-calculus is a formalism for expressing functional programs.
It includes variables, function application, and the ability to create new functions using ...

TODO: summarize lambda calculus and types in one paragraph

\noindent\textbf{Definition: $\mathcal{D}$.}
A DSL $\mathcal{D}$ is a set of typed $\lambda$-calculus expressions.
\\\noindent\textbf{Definition: $\theta$.}
A weight vector $\theta$ for a DSL $\mathcal{D}$ is a vector of $|\mathcal{D}| + 1$ real numbers:
one number for each DSL primitive $e:\tau\in \mathcal{D}$, written $\theta_e$,
and a weight controlling the probability of a variable occurring in a program, written $\theta_{\text{var}}$.

Algorithm~\ref{programGenerativeModel} is a procedure for drawing
samples from $\probability[p|\mathcal{D},\theta]$.  In practice, we
enumerate programs rather than sampling them.  Enumeration proceeds by
a depth-first search over the random choices made by
Algorithm~\ref{programGenerativeModel}; we wrap the depth-first search
in iterative deepening to (approximately) build $\lambda$-calculus
expressions in order of their probability.

\begin{algorithm}[tb]
   \caption{Generative model over programs}
   \label{programGenerativeModel}
   \begin{algorithmic}
     \STATE \textbf{function} sampleProgramFromDSL$(\mathcal{D}, \theta, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \STATE \textbf{return} sample$(\mathcal{D}, \theta, \varnothing, \tau)$
\STATE
     \STATE \textbf{function} sample$(\mathcal{D}, \theta, \mathcal{E}, \tau)$:
  \STATE {\bfseries Input:} DSL $\mathcal{D}$, weight vector $\theta$, environment $\mathcal{E}$, type $\tau$
  \STATE \textbf{Output:} a program whose type unifies with $\tau$
  \IF{$\tau = \alpha\to\beta$}
  \STATE var $\gets$ an unused variable name
  \STATE body $\sim$ sample$(\mathcal{D},\theta,\{\text{var}:\alpha\}\cup\mathcal{E},\beta)$
   \STATE \textbf{return} $\lambda \text{var}.$ body
   \ENDIF
   %   \ELSE
   \STATE $\text{primitives} \gets\{p | &p:\alpha\to\cdots\to\beta \in \mathcal{D}\cup\mathcal{E}$
   \STATE \hspace{2.5cm}$\text{if canUnify}(\tau,\beta) \} $

   \STATE Draw $e\sim \text{primitives}$, w.p. $\propto\theta_e$ if $e\in \mathcal{D}$
   \STATE \hspace{3.1cm}w.p. $\propto\frac{\theta_{var}}{|\text{variables}|}$ if $e\in \mathcal{E}$
   \STATE Let $e:\alpha_1\to\alpha_2\to\cdots\to \alpha_K\to\beta$. Unify $\tau$ with $\beta$.
%   \STATE unify$(\tau,\beta)$
   \FOR{$k=1$ {\bfseries to} $K$}
 \STATE $a_k\sim\text{sample}(\mathcal{D},\theta,\mathcal{E},\alpha_k)$
 \ENDFOR
 \STATE \textbf{return} $e\;a_1\; a_2\; \cdots\; a_K$
 %  \UNTIL{$noChange$ is $true$}
\end{algorithmic}
\end{algorithm}

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:% for using the simple strategy of enumerating programs:
\begin{itemize}
\item Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.
\item A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the
  search algorithm is very simple.  
\end{itemize}
A main drawback of an enumerative search algorithm is that we have no
efficient means of solving for arbitrary constants that might occur in the
program. In Section~\ref{regressionSection},
we will show how to find programs with real-valued constants
by automatically differentiating through the program and setting the constants using gradient descent.

@@

\begin{algorithm}[tb]
   \caption{Grammar Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $\log \probability[\mathcal{D}]\stackrel{+}{ = } -\lambda\sum_{p\in \mathcal{D}} \text{size}(p)$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ a subtree of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \IF{$\text{score}(\mathcal{D}') > \text{score}(\mathcal{D})$}
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ELSE
     \STATE\textbf{return} $\mathcal{D},\theta^*(\mathcal{D})$
     \ENDIF
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}
\section{Program Representation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.

@@

\subsection{Symbolic Regression}\label{regressionSection}
We show how to use \system to infer programs containing both discrete
structure and continuous parameters. The high-level idea is to synthesize programs with unspecified-real-valued parameters, and to fit those parameters using gradient descent.
Concretely, we ask the algorithm to
solve a set of 1000 symbolic regression problems, each a polynomial of
degree 0, 1, or 2, where our observations $x$ take the form of $N$
input/output examples, which we write as $x = \left\{(i_n,o_n)
\right\}_{n\leq N}$. For example, one task is to infer a program
calculating $3x + 2$, and the observations are the input-output
examples $\left\{(-1,-1),(0,2),(1,5) \right\}$.

@@
